---
title: "Lab 10. Dimensionality reduction. PCA"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
library(tidyverse)
library(ggfortify)
#Sys.setlocale(locale = "ru_RU.UTF-8")
```

## Principal component analysis (PCA)

### 1. Main problem
Sometimes you have a huge amount of variables. So, to make your data profitable you need to reduce number of variables saving without losing the precious information. 

* Principal component analysis (PCA)
* Linear discriminant analysis (LDA)
* Multidimensional scaling (MDS)
* ...

### 2. Data
This is a dataset from [Huttenlocher, Vasilyeva, Cymerman, Levine 2002]. The authors analysed 46 pairs of mothers and children (aged from 47 to 59 months, mean age -- 54). They recorded and transcribed 2 hours from each child per day. In their study, they compared the number of noun phrases per utterance in mother speech to the number of noun phrases per utterance in child speech.
```{r}
df <- read_csv("https://raw.githubusercontent.com/agricolamz/r_on_line_course_data/master/Huttenlocher_2001.csv")

df %>%
  ggplot(aes(mother, child))+
  geom_point(color = "darkgreen", size = 3)+
  stat_ellipse(linetype=2)+
  theme_bw()
```

### 3. PCA
PCA is essentially a rotation of the coordinate axes, chosen such that each successful axis captures as much variance as possible. We can reduce 2 dementions to one using a regression:

```{r}
fit <- lm(child~mother, data = df)
df$model <- predict(fit)

df %>%
  ggplot(aes(mother, child))+
  geom_line(aes(mother, model), color = "blue")+
  geom_point(color = "darkgreen", size = 3)+
  stat_ellipse(linetype=2)+
  theme_bw()
```

We used regression for predicting value of one variable by another variable.

```{r}

df %>%
 ggplot(aes(mother, child))+
stat_ellipse(linetype=2)+
geom_segment(aes(x=min(mother), y=1.8, xend=2, yend=1.8), size=0.5, color = "red", arrow =  arrow(angle = 10, type = "closed", ends = "first"))+
geom_segment(aes(x=2, y=min(child), xend=2, yend=1.8), size=0.5, color = "red", arrow =  arrow(angle = 10, type = "closed"))+
geom_line(aes(mother, model), color = "blue")+
geom_point(color = "darkgreen", size = 3)+
scale_y_continuous(breaks = c(1.2, 1.4, 1.6, 1.8, 2.0))+
 theme_bw()+
theme(axis.text.x = element_text(color=c("black", "black", "black", "red", "black"), size=c(9, 9, 9, 14, 9)),
      axis.text.y = element_text(color=c("black", "black", "black", "red", "black", "black"), size=c(9, 9, 9, 14, 9, 9)))
```

In PCA we change coordinate system and start predicting variables' values using less variables.

```{r}
pca <- prcomp(df)
pca$rotation
pca$center
```

The number of the PCs is always equal to the number of variables. 

The main point of PCA is that if cumulative proportion of explained variance is high we can drop some PCs. So, we need know the following things:

* What is the cumulative proportion of explained variance?
```{r, echo = TRUE}
summary(prcomp(df))
```

So, PC1 explains only 78.9% of the variance in our data.

* How PCs are rotated comparing to the old axes?
```{r, echo = TRUE}
#df <- read.csv("https://raw.githubusercontent.com/agricolamz/r_on_line_course_data/master/Huttenlocher_2001.csv")
prcomp(df)
```

So the formula for the first component rotation is
$$PC1 = 0.6724959 \times child + 0.7401009  \times mother$$
The formula for the second component rotation is
$$PC2 = -0.7401009 \times child + 0.6724959  \times mother$$

From now we can change the axes. We use the `autoplot()` function from `ggfortify` package to produce the graph:
```{r}
autoplot(pca,
         loadings = TRUE,
         loadings.label = TRUE)+
  theme_bw()+
  stat_ellipse(linetype=2)
```


### Summary:

* If the cumulative proportion of explained variance for some PCs is high, we can change coordinate system and start predicting variables' values using less variables.
* We can even make a regresion or clusterisation model.
* PCA for categorical variables is called Multiple correspondence analysis (MCA)

### R functions

There are several functions for PCA, MCA and their visualisation.

* PCA: prcomp()
* PCA: princomp()
* PCA: FactoMineR::PCA()
* PCA: ade4::dudi.pca()
* PCA: amap::acp()
* PCA visualisation: ggfortify::autoplot


### 2 Gospels' frequency word lists 

The gospels of Matthew, Mark, and Luke are referred to as the Synoptic Gospels and stand in contrast to John, whose content is comparatively distinct. This dataset (https://tinyurl.com/y8tcf3uw) contains frequency of selected words (without stopwords, without pronouns and without frequent word "Jesus") as attested in four gospels of the New Testament.

For some visualisations you will need assign row names to the dataframe:

```{r}
gospels <- read.csv("https://tinyurl.com/y8tcf3uw")
row.names(gospels) <- gospels$word
```

#### 1.2 Apply PCA to four continuous variables. Use `prcomp()` function. What is the cumulative proportion of explained variance for the first and second component?

```{r}
PCA <- prcomp(gospels[,2:5], center = TRUE, scale. = TRUE)
summary(PCA)
```

#### 2.2 Use the `autoplot()` function of the library ggfortify for creating plot like this. 
See more examples here: https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_pca.html

```{r}
autoplot(PCA,
         shape = FALSE,
         loadings = TRUE,
         label = TRUE,
         loadings.label = TRUE)+
  theme_bw()
```

#### 2.3 Predict the coordinates for the word "Jesus", which have the following frequencies: John = 0.05, Luke = 0.01, Mark = 0.02, Matthew = 0.02.

```{r}
predict(PCA, data.frame(John = 0.05, Luke = 0.01, Mark = 0.02, Matthew = 0.02))
```

We can also look at PC2 and PC3 components:
```{r}
autoplot(PCA, x=2, y=3,
         shape = FALSE,
         loadings = TRUE,
         label = TRUE,
         loadings.label = TRUE)+
  theme_bw()
```

### Useful links

* FactoMineR for PCA [link](http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/)
